Functional pant indicators - wetland, indicator documentation

### 2023-04-13 DRAFT_functional plant indicators_wetland.Rmd
fpi wetland - until analysis of results

---
title: "Template for indicator documentation"
output:
  html_document
---

@@

Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
Name te file DRAFT_myIndicatorName.Rmd.

@@

<!-- Replace 'indicator name' with yor own indicator name  -->
# Functional plant indicators (Moisture, Light, pH, Nitrogen)

@@

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(tidyr)
library(dplyr)
library(plyr)
library(stringr)
library(tidyverse)
library(readxl)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "våtmark"
Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
ECT       <- "" # e.g. "Structural state characteristic"
Contact   <- "Joachim Töpper"
```

@@

```{r}
st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                   layer="ANO_Art")
ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                  layer="ANO_SurveyPoint")
head(ANO.sp)
head(ANO.geo)
```

Plant indicators from Tyler et al. 2021
```{r}
## Tyler indicator data
ind.dat <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
                        sep = '\t', header=T, quote = '')
head(ind.datr)

@@

Generalized species lists NiN
```{r}
load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators//reference from NiN/Eco_State.RData")
str(Eco_State)
```

#### Plant indicator data - data handling
```{r}

names(ind.dat)
names(ind.dat)[1] <- 'species'
ind.dat$species <- as.factor(ind.dat$species)
summary(ind.dat$species)
ind.dat <- ind.dat[!is.na(ind.dat$species),] # removing species-NAs
ind.dat[,'species.orig'] <- ind.dat[,'species'] # make a backup of the original species variable
ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2) # trimming away sub-species & co, and descriptor info

@@

#### ANO monitoring data - data handling
```{r}
head(ANO.sp)
head(ANO.geo)

@@

#### NiN reference data - data handling

```{r}

@@

## matching with NiN ecosystem types
# NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[19:23])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
NiN.wetland[1,]
names(NiN.wetland)

#NiN.wetland <- NiN.wetland[,c(1:41,46:48,59:69)]
cbind(colnames(NiN.wetland),
      c("",
        'V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
        'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
        'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
        'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
        'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
        'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
        'V3-C2','V1-C5',
        'V1-C6a','V1-C6b',
        'V1-C7a','V1-C7b',
        'V1-C8a','V1-C8b',
        'V2-C1a','V2-C1b',
        'V2-C2a','V2-C2b',
        'V2-C3a','V2-C3b',
        "V4-C4","V4-C5",
        "V4-C2","V4-C3",
        'V8-C1','V8-C2','V8-C3',
        rep("",10),
        rep("",5)
      )
)

NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:63)]
colnames(NiN.wetland)[2:46] <- c('V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
                              'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
                              'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
                              'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
                              'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
                              'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
                              'V3-C2','V1-C5',
                              'V1-C6a','V1-C6b',
                              'V1-C7a','V1-C7b',
                              'V1-C8a','V1-C8b',
                              'V2-C1a','V2-C1b',
                              'V2-C2a','V2-C2b',
                              'V2-C3a','V2-C3b',
                              'V4-C2','V4-C3',
                              'V8-C1','V8-C2','V8-C3'
)
head(NiN.wetland)

@@

summary(NiN.wetland)
summary(NiN.wetland.cov)

@@

>>>>

---
title: "Functional pant indicators - wetland, indicator documentation"
output:
  html_document
---

@@

Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
Name the file DRAFT_myIndicatorName.Rmd.

@@

<!-- Replace 'indicator name' with your own indicator name  -->
# Functional plant indicators (Moisture, Light, pH, Nitrogen)

@@

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(tidyr)
library(dplyr)
library(plyr)
library(stringr)
library(tidyverse)
library(readxl)
library(broom)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(ggplot2)
library(tmap)
library(raster)
library(stars)

library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "våtmark"
Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
ECT       <- "Functional state characteristic"
Contact   <- "Joachim Töpper"
```

@@

```{r, echo = F}
st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                   layer="ANO_Art")
ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                  layer="ANO_SurveyPoint")
head(ANO.sp)
head(ANO.geo)
```

Plant indicators from Tyler et al. 2021
```{r, echo = F}
## Tyler indicator data
ind.dat <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
                        sep = '\t', header=T, quote = '')
head(ind.dat)

@@

Generalized species lists NiN
```{r, echo = F}
load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators//reference from NiN/Eco_State.RData")
str(Eco_State)
```

#### Data handling
- Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)

```{r, include = F}

### Plant indicator data
names(ind.dat)
names(ind.dat)[1] <- 'species'
ind.dat$species <- as.factor(ind.dat$species)
summary(ind.dat$species)
ind.dat <- ind.dat[!is.na(ind.dat$species),] # removing species-NAs
ind.dat[,'species.orig'] <- ind.dat[,'species'] # make a backup of the original species variable
ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2) # trimming away sub-species & co, and descriptor info

@@

### ANO monitoring data

head(ANO.sp)
head(ANO.geo)

@@

### NiN reference data - data handling

@@

## matching with NiN ecosystem types - for wetlands
# NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[15:18])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
NiN.wetland[1,]
names(NiN.wetland)

cbind(colnames(NiN.wetland),
      c("",
        'V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
        'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
        'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
        'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
        'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
        'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
        'V3-C2','V1-C5',
        'V1-C6a','V1-C6b',
        'V1-C7a','V1-C7b',
        'V1-C8a','V1-C8b',
        'V2-C1a','V2-C1b',
        'V2-C2a','V2-C2b',
        'V2-C3a','V2-C3b',
        "V4-C2","V4-C3",
        "","",
        'V8-C1','V8-C2','V8-C3',
        rep("",10),
        rep("",4)
      )
)

NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:62)]
colnames(NiN.wetland)[2:46] <- c('V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
                              'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
                              'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
                              'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
                              'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
                              'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
                              'V3-C2','V1-C5',
                              'V1-C6a','V1-C6b',
                              'V1-C7a','V1-C7b',
                              'V1-C8a','V1-C8b',
                              'V2-C1a','V2-C1b',
                              'V2-C2a','V2-C2b',
                              'V2-C3a','V2-C3b',
                              'V4-C2','V4-C3',
                              'V8-C1','V8-C2','V8-C3'
)
head(NiN.wetland)

@@

NiN.wetland.cov$sp <- as.factor(NiN.wetland.cov$sp)
```
leaving us with the monitoring data including plant indicators (ANO.sp.ind) and the reference data including plant indicators (NiN.wetland.cov)

```{r}
#summary(NiN.wetland)
head(ANO.sp.ind)
head(NiN.wetland.cov)
```
For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
For wetland ecosystem we are testing "Light", "Moisture", "Soil_reaction_pH", and "Nitrogen". In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.

##### bootstrap function for frequency abundance
- function to calculate community weighted means of selected indicator values (ind)
- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
- with a given number of iterations (iter),
- with species given a certain minimum abundance ocurring in all bootstraps (obl), and
- with a given resampling ratio of the original species list (rat)
- in every bootstrap iteration the abundance of the sampled species can be 
- randomly changed by +/-1 if wished (var.abun)

```{r, echo = F}
indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {

  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
  colnames(ind.b) <- colnames(abun)
  ind.b <- as.data.frame(ind.b)

  ind <- as.data.frame(ind)
  ind.list <- as.list(1:length(colnames(ind)))
  names(ind.list) <- colnames(ind)

  for (k in 1:length(colnames(ind)) ) {
    ind.list[[k]] <- ind.b }

  for (j in 1:length(colnames(abun)) ) {

    dat <- cbind(sp,abun[,j],ind)
    dat <- dat[dat[,2]>0,]            # only species that are present in the ecoystem
    dat <- dat[!is.na(dat[,3]),]      # only species that have Ellenberg values

    for (i in 1:iter) {

      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2]>=obl])) *rat,0), replace=F)  
      dat.b <- rbind(dat[dat[,2] >= obl,],
                     dat[match(speciesSample,dat$sp),]
      )

      if (var.abun==T) {
        for (m in 1:nrow(coverscale[-1,]) ) {
          xxx <- dat.b[dat.b[,2]==coverscale[-1,][m,2],2]
          if ( m==1 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.5, 0.5, 0.0, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==2 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.2, 0.3, 0.5, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==3 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.2, 0.3, 0.5, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==4 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.2, 0.3, 0.5, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==5 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.2, 0.3, 0.5, 0.0) ,size=length(xxx),replace=T) }
          if ( m==6 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.0, 0.2, 0.3, 0.5) ,size=length(xxx),replace=T) }
        }
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]<=(0),2] <- 0.01
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]>1,2] <- 1
      }

      for (k in 1:length(colnames(ind))) {

        if ( nrow(dat.b)>2 ) {

          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
          ind.list[[k]][i,j] <- ind.b

        } else {ind.list[[k]][i,j] <- NA}

      }

#      print(paste(i,"",j)) 
    }

  }
  return(ind.list)
}
```

Running the bootstraps

```{r, echo = T, eval = F}
colnames(NiN.wetland)
# 1st column is the species
# 2nd-46th column is the abundances of sp in different ecosystem types
# 47th-51st column is the indicator values of the respective sp
# we choose 1000 iterations
# species with abundance 1 (i.e. 100%, must be included in each sample)
# each sample resamples 2/3 of the number of species
# the abundance of the resampled species may vary (see bootstrap function for details)
wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
                          iter=1000,obl=1,rat=2/3,var.abun=T)

# fixing NaNs
for (i in 1:length(wetland.ref.cov) ) {
  for (j in 1:ncol(wetland.ref.cov[[i]]) ) {
    v <- wetland.ref.cov[[i]][,j]
    v[is.nan(v)] <- NA
    wetland.ref.cov[[i]][,j] <- v
  }
}
```

```{r, include = F}
load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_wetland.RData")
```

```{r}
head(wetland.ref.cov[[1]])
```

This results in a list with a slot for every selected indicator, and in every slot there's as many columns as there are NiN species lists and as many row as there were iterations in the bootstrap.
Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here we define things in the following way:

Median = reference values
0.025 and 0.975 quantiles = lower and upper limit values
min and max of the respective indicator's scale = min/max values

```{r, echo = F}

# every NiN-type is represented by one 'generalisert artsliste'
# some NiN-types are represented by two such species lists
# in some cases two NiN-types are represented by the same species list
head(wetland.ref.cov[[1]])
wetland.ref.cov[[1]][0,]

# checking the actual NiN-types in the wetland lists
wetland.NiNtypes <- colnames(wetland.ref.cov[["Light"]])
wetland.NiNtypes <- substr(wetland.NiNtypes,1,5)
unique(wetland.NiNtypes)

# 4 indicator-value indicators: Tyler's Light, Moisture, Soil_reaction_pH, "Nitrogen"
indEll.n=4
# creating a table to hold:
# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
# for every nature type (nrows)
tab <- matrix(ncol=3*indEll.n, nrow=length(unique(wetland.NiNtypes)) ) # 43 basic ecosystem types
# coercing the values into the table
# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
names(wetland.ref.cov[["Light"]])
x <- c(27,28,41:45)

for (i in 1:length(x) ) {
  tab[i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
}

tab <- as.data.frame(tab)
tab$NiN <- NA
tab$NiN[1:length(x)] <- names(wetland.ref.cov[[1]])[x]

# NiN-types represented by several species lists
wetland.NiNtypes2 <- wetland.NiNtypes[-x]
unique(wetland.NiNtypes2)
grep(pattern=unique(wetland.NiNtypes2)[1], x=wetland.NiNtypes) # finds columns in e.g. colnames(wetland.ref.cov[["Continentality"]]) that match the first NiN-type

for (i in 1:length(unique(wetland.NiNtypes2)) ) {
  tab[length(x)+i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[length(x)+i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[length(x)+i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[length(x)+i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab$NiN[length(x)+i] <- unique(wetland.NiNtypes2)[i]

}

tab

# making it a proper data frame
round(tab[,1:12],digits=2)

colnames(tab) <- c("Light_q2.5","Light_q50","Light_q97.5",
                   "Moist_q2.5","Moist_q50","Moist_q97.5",
                   "pH_q2.5","pH_q50","pH_q97.5",
                   "Nitrogen_q2.5","Nitrogen_q50","Nitrogen_q97.5",
                   "NiN")
summary(tab)
tab$NiN <- gsub("C", "C-", tab$NiN) # add extra hyphon after C for NiN-types
tab

# restructuring into separate indicators for lower (q2.5) and higher (q97.5) than reference value (=median, q50)
y.Light <- numeric(length=nrow(tab)*2)
y.Light[((1:dim(tab)[1])*2)-1] <- tab$Light_q2.5 
y.Light[((1:dim(tab)[1])*2)] <- tab$Light_q97.5

y.Moist <- numeric(length=nrow(tab)*2)
y.Moist[((1:dim(tab)[1])*2)-1] <- tab$Moist_q2.5 
y.Moist[((1:dim(tab)[1])*2)] <- tab$Moist_q97.5

y.pH <- numeric(length=nrow(tab)*2)
y.pH[((1:dim(tab)[1])*2)-1] <- tab$pH_q2.5 
y.pH[((1:dim(tab)[1])*2)] <- tab$pH_q97.5

y.Nitrogen <- numeric(length=nrow(tab)*2)
y.Nitrogen[((1:dim(tab)[1])*2)-1] <- tab$Nitrogen_q2.5 
y.Nitrogen[((1:dim(tab)[1])*2)] <- tab$Nitrogen_q97.5

# creating final objects holding the reference and limit values for all indicators

# ref object for indicators
wetland.ref.cov.val <- data.frame(N1=rep('wetland',(nrow(tab)*2*indEll.n)),
                              hoved=c(rep('NA',(nrow(tab)*2*indEll.n))),
                              grunn=c(rep(rep(tab$NiN,each=2),indEll.n)),
                              county=rep('all',(nrow(tab)*2*indEll.n)),
                              region=rep('all',(nrow(tab)*2*indEll.n)),
                              Ind=c(rep(c('Light1','Light2'),nrow(tab)),
                                    rep(c('Moist1','Moist2'),nrow(tab)),
                                    rep(c('pH1','pH2'),nrow(tab)),
                                    rep(c('Nitrogen1','Nitrogen2'),nrow(tab))
                              ),
                              Rv=c(rep(tab$Light_q50,each=2),
                                   rep(tab$Moist_q50,each=2),
                                   rep(tab$pH_q50,each=2),
                                   rep(tab$Nitrogen_q50,each=2)
                              ),
                              Gv=c(y.Light,y.Moist,y.pH,y.Nitrogen),
                              maxmin=c(rep(c(1,7),nrow(tab)), # 7 levels of light
                                       rep(c(1,12),nrow(tab)), # 12 levels of moisture
                                       rep(c(1,8),nrow(tab)), # 8 levels of soil reaction pH
                                       rep(c(1,9),nrow(tab))  # 9 levels of nitrogen
                              )
)

wetland.ref.cov.val
wetland.ref.cov.val$grunn <- as.factor(wetland.ref.cov.val$grunn)
wetland.ref.cov.val$Ind <- as.factor(wetland.ref.cov.val$Ind)
summary(wetland.ref.cov.val)

@@

```{r}
head(wetland.ref.cov.val)
```

Once testing data (ANO), and reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO-community data and scale them against the scaling values from the reference distribution.

Here is the scaling function (with and without truncation at 1)
```{r}

#### scaled values ####
r.s <- 1    # reference value
l.s <- 0.6  # limit value
a.s <- 0    # abscence of indicator, or indicator at maximum

#### function for calculating scaled values for measured value ####

## scaling function including truncation
scal <- function() {
  # place to hold the result
  x <- numeric()
  if (maxmin < ref) {
    # values >= the reference value equal 1
    if (val >= ref) {x <- 1}
    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
    # value equals or lower than maxmin
    if (val <= maxmin) {x <-0}
  } else {
    # values <= the reference value equal 1
    if (val <= ref) {x <- 1}
    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
    # value equals or larger than maxmin
    if (val >= maxmin) {x <-0}
  }
  return(x)

}

## scaling function without truncation
scal.2 <- function() {
  # place to hold the result
  x <- numeric()
  if (maxmin < ref) {
    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
    # value equal or lower than maxmin
    if (val <= maxmin) {x <-0}
  } else {
    # values <= the reference value estimated from the linear relationship for lim < x < ref (line below)
    if (val <= ref) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
    if (val > lim & val < maxmin) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
    # value equal og larger than maxmin
    if (val >= maxmin) {x <-0}
  }
  return(x)

}
```

We then can prepare a list of dataframes to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
```{r, echo = T, eval = F}

unique(ANO.geo$hovedtype_rute) # NiN types in data
unique(substr(wetland.ref.cov.val$grunn,1,2)) # NiN types in reference
#### creating dataframe to hold the results for wetlands ####
# all ANO points
nrow(ANO.geo)
# all wetland ANO points
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),])
# all wetland ANO points with a NiN-type represented in the reference
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(wetland.ref.cov.val$grunn,1,2)),])
# ok, we'll be losing 70 (our of 1349) that are not covered by the reference
ANO.wetland <- ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),]

head(ANO.wetland)
# update row-numbers
row.names(ANO.wetland) <- 1:nrow(ANO.wetland)
head(ANO.wetland)
dim(ANO.wetland)
colnames(ANO.wetland)

length(levels(as.factor(ANO.wetland$ano_flate_id)))
length(levels(as.factor(ANO.wetland$ano_punkt_id)))
summary(as.factor(ANO.wetland$ano_punkt_id))
# four points that are double
ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0159_55",] # double registration, said so in comment. -> choose row 207 over 206
ANO.wetland <- ANO.wetland[-206,]
row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0283_22",] # 2019 & 2021. Lot of NA's in 2019 -> omit 2019
ANO.wetland <- ANO.wetland[-156,]
row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0363_24",]
ANO.wetland[ANO.wetland$ano_flate_id=="ANO0363","ano_punkt_id"] # point-ID 15 is missing, but 24 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both, call the second obs the one that's missing
ANO.wetland[311,"ano_punkt_id"] <- "ANO0363_15"
ANO.wetland[ANO.wetland$ano_punkt_id=="ANO1550_64",] # point-ID 66 is missing, but 64 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both
ANO.wetland[ANO.wetland$ano_flate_id=="ANO1550","ano_punkt_id"] # point-ID 66 is missing, but 64 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both
ANO.wetland[1273,"ano_punkt_id"] <- "ANO1550_66"

unique(ANO.wetland$hovedoekosystem_punkt)
unique(ANO.wetland$hovedtype_rute)
unique(ANO.wetland$kartleggingsenhet_1m2)
ANO.wetland$hovedtype_rute <- factor(ANO.wetland$hovedtype_rute)
ANO.wetland$kartleggingsenhet_1m2 <- factor(ANO.wetland$kartleggingsenhet_1m2)
summary(ANO.wetland$Hovedtype_rute)
summary(ANO.wetland$Kartleggingsenhet_rute)

results.wetland <- list()
ind <- unique(wetland.ref.cov.val$Ind)
# choose columns for site description
colnames(ANO.wetland)
results.wetland[['original']] <- ANO.wetland
# drop geometry
st_geometry(results.wetland[['original']]) <- NULL
results.wetland[['original']]

# add columns for indicators
nvar.site <- ncol(results.wetland[['original']])
for (i in 1:length(ind) ) {results.wetland[['original']][,i+nvar.site] <- NA}
colnames(results.wetland[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.wetland[['original']][,i] <- as.numeric(results.wetland[['original']][,i])}
summary(results.wetland[['original']])
#results.wetland[['original']]$Region <- as.factor(results.wetland[['original']]$Region)
results.wetland[['original']]$GlobalID <- as.factor(results.wetland[['original']]$GlobalID)
results.wetland[['original']]$ano_flate_id <- as.factor(results.wetland[['original']]$ano_flate_id)
results.wetland[['original']]$ano_punkt_id <- as.factor(results.wetland[['original']]$ano_punkt_id)
results.wetland[['original']]$hovedoekosystem_punkt <- as.factor(results.wetland[['original']]$hovedoekosystem_punkt)
#results.wetland[['original']]$Hovedoekosystem_rute  <- as.factor(results.wetland[['original']]$Hovedoekosystem_rute )
results.wetland[['original']]$kartleggingsenhet_1m2 <- as.factor(results.wetland[['original']]$kartleggingsenhet_1m2)
results.wetland[['original']]$hovedtype_rute    <- as.factor(results.wetland[['original']]$hovedtype_rute)

# roll out
results.wetland[['scaled']] <- results.wetland[['non-truncated']] <- results.wetland[['original']]

#### calculating scaled and non-truncated values for the indicators based on the dataset ####
for (i in 1:nrow(ANO.wetland) ) {  #
  tryCatch({
    print(i)
    print(paste(ANO.wetland$ano_flate_id[i]))
    print(paste(ANO.wetland$ano_punkt_id[i]))
#    ANO.wetland$Hovedoekosystem_sirkel[i]
#    ANO.wetland$Hovedoekosystem_rute[i]

    # if the ANO.hovedtype exists in the reference
    if (ANO.wetland$hovedtype_rute[i] %in% unique(substr(wetland.ref.cov.val$grunn,1,2)) ) {

      # if there is any species present in current ANO point  
      if ( length(ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),'Species']) > 0 ) {

          # Light
          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Light')]
          results.wetland[['original']][i,'richness'] <- nrow(dat)
          dat <- dat[!is.na(dat$Light),]

          if ( nrow(dat)>0 ) {

            val <- sum(dat[,'art_dekning'] * dat[,'Light'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
            # lower part of distribution
            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.wetland dataframe
            results.wetland[['scaled']][i,'Light1'] <- scal() 
            results.wetland[['non-truncated']][i,'Light1'] <- scal.2() 
            results.wetland[['original']][i,'Light1'] <- val

            # upper part of distribution
            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.wetland dataframe
            results.wetland[['scaled']][i,'Light2'] <- scal() 
            results.wetland[['non-truncated']][i,'Light2'] <- scal.2() 
            results.wetland[['original']][i,'Light2'] <- val
          }
